Experience Report: Haskell in Computational Biology

### 2012-03-07 mrfy_experience_report.tex
a few notes and a cut at the argument in the intro

Computational biologists detect homologies by building 
algorithms which, given an amino-acid sequence,\remark{This sequence
  (the sequence ``under test'') needs a name.  (Urk! I see below it is
  called the ``query sequence.''  This name needs to come in earlier.) ---NR}
find known proteins of
similar structure.
When the similar proteins are formed from amino-acid sequences that
are nearly identical to the sequence under test, the homologies can be
detected by  
simple string-matching algorithms.\remark{Can we just drop the
  sentence about string matching?}
When the similar proteins are formed from amino-acid sequences that
are not too far in edit distance from the sequence under test, the homologies can be
detected by
a family of algorithms called 
\textit{hidden Markov models}.\remark{citation needed?}
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---the MRFy algorithm---can detect homologies
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.

@@

@@

The protein alignment is used to train a \emph{hidden-Markov model}.
A~hidden-Markov model is a probabilistic finite-state machine which 
can assign a probability to any amino-acid sequence under test.
\remark{What is the biological or computational meaning of the
  probability?}
Proteins whose sequences have higher probabilities are more likely to
be structurally related to the proteins in the alignment.
\remark{Should we say ``homologous'' rather than  ``structurally
  related''?
  If~so, where should this jargon be introduced?}
The states of the hidden-Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
For a given column~$C$ of the alignment, the hidden-Markov model has a
\emph{match state}.
The match state determines the probability that the sequence under
test has a particular amino acid in column~$C$.
The match state assigns high probability to the amino acids that
appear in column~$C$, and a low but nonzero probability to other amino
acids.
\item
For every column of the alignment, the hidden-Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the sequence under
test has \emph{no} amino acid  in column~$C$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$C$,
and low probability if there are few gaps in column~$C$.
\item
Between every adjacent pair of columns in the alignment,\remark{This isn't right}
the hidden-Markov model has an
\emph{insertion state}.
The insertion state determines the probability that the sequence under
test has an ``extra'' amino acid that does not correspond to any
column in the alignment.
The insertion state assigns high probability to \emph{[some amino
    acids]}, and low probability to \emph{[other amino acids]}.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden-Markov model also has distinguished \emph{begin} and \emph{end} states.
\figref{plan7} shows a hidden-Markov model derived from an alignment
with just four columns.
\emph{[I'm sure I don't have the proper explanation of the probabilities.]}
\remark{LOST: Each edge in the 
graph has a weight which is the probability of the associated state transition.}
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.
What looks like an 
intimidating geometry is thus completely explained by the fact that we have 
insertions, deletions, and substitutions due to evolution.
\remark{The evolution seems like a distraction to me here, and we hit
  it above.  Can we simply drop this sentence?}
\emph{[HOW ABOUT SOME HASKELL TYPE DEFINITIONS AT OR NEAR THIS POINT?
    I THINK WE'RE READY.]}

>>>>

Computational biologists detect homologies by building 
algorithms which, given an amino-acid sequence,\remark{This sequence
  (the sequence ``under test'') needs a name.  (Urk! I see below it is
  called the ``query sequence.''  This name needs to come in earlier.) ---NR}
find known proteins of
similar structure.
When the similar proteins are formed from amino-acid sequences that
are nearly identical to the sequence under test, the homologies can be
detected by  
simple string-matching algorithms.\remark{Can we just drop the
  sentence about string matching?}
When the similar proteins are formed from amino-acid sequences that
are not too far in edit distance from the sequence under test, the homologies can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{hmmer}.\remark{citation needed?}
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---the MRFy algorithm---can detect homologies
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.

@@

We present an explanation of the algorithms relevant to MRFy, because there 
exists a natural fit between the mathematics behind the algorithms and the 
Haskell implementation of those algorithms.
If the reader understands the 
algorithms, he or she is more likely to appreciate the benefits of implementing 
them in Haskell.
We will first present the more traditional algorithm used in 
hidden Markov models, followed by our variation on it and the stochastic search 
approach.
We conclude that implementations of mathematically-intensive 
algorithms such as these lend themselves to the Haskell language, particularly 
when the programmer needs a flexible implementation that simplifies 
experimenting with variants of the algorithm.

@@

The protein alignment is used to train a \emph{hidden-Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden-Markov model is a probabilistic finite-state machine which 
can assign a probability to any amino-acid sequence under test.
\remark{What is the biological or computational meaning of the
  probability?}
Proteins whose sequences have higher probabilities are more likely to
be structurally related to the proteins in the alignment.
\remark{Should we say ``homologous'' rather than  ``structurally
  related''?
  If~so, where should this jargon be introduced?}
The states of the hidden-Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
For a given column~$C$ of the alignment, the hidden-Markov model has a
\emph{match state}.
The match state determines the probability that the sequence under
test has a particular amino acid in column~$C$.
The match state assigns high probability to the amino acids that
appear in column~$C$, and a low but nonzero probability to other amino
acids.
\item
For every column of the alignment, the hidden-Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the sequence under
test has \emph{no} amino acid  in column~$C$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$C$,
and low probability if there are few gaps in column~$C$.
\item
Between every adjacent pair of columns in the alignment,\remark{This isn't right}
the hidden-Markov model has an
\emph{insertion state}.\remark{concept of consensus columns is what's missing}
The insertion state determines the probability that the sequence under
test has an ``extra'' amino acid that does not correspond to any
column in the alignment.
The insertion state assigns high probability to \emph{[some amino
    acids]}, and low probability to \emph{[other amino acids]}.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden-Markov model also has distinguished \emph{begin} and \emph{end} states.
\figref{plan7} shows a hidden-Markov model derived from an alignment
with just four columns.
\emph{[I'm sure I don't have the proper explanation of the probabilities.]}
\remark{LOST: Each edge in the 
graph has a weight which is the probability of the associated state transition.}
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.
What looks like an 
intimidating geometry is thus completely explained by the fact that we have 
insertions, deletions, and substitutions due to evolution.
\remark{The evolution seems like a distraction to me here, and we hit
  it above.  Can we simply drop this sentence?}
\emph{[HOW ABOUT SOME HASKELL TYPE DEFINITIONS AT OR NEAR THIS POINT?
    I THINK WE'RE READY.]}
